---
title: "Differential Expression Analysis"
author: "Jack Gisby"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 7
---

# Load data and code
## Load packages and custom functions

Custom functions for loading data, running differential expression and 
enrichment analyses, and making plots can be found in the `../R` directory. The 
main functions and their arguments are described by comments within these files. 

```{r load_functions, message=FALSE, warning=FALSE}

# load necessary packages
library(data.table)
library(SummarizedExperiment)
library(variancePartition)
library(BiocParallel)
library(edgeR)
library(lmerTest)
library(GSVA)
library(biomaRt)

# load custom functions from the R/ directory
source("../R/preprocessing.R")
source("../R/de.R")

set.seed(1)

```

## Load RNA-seq data

The raw RNA-seq counts are loaded, transformed into `SummarizedExperiment` objects,
and split into the "Wave 1" and "Wave 2" cohorts. `SnowParam` is also initialised
for the differential expression analysis using the `variancePartition` package.

```{r load_rna_data, message=FALSE, warning=FALSE}

# get the raw RNA-seq counts
htseq_counts <- data.frame(fread("../data/htseq_counts.csv"))

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

# get the wave 1 and wave 2 counts and their associated metadata
wave1_counts <- get_summarized_experiment(
  htseq_counts, 
  "../data/w1_metadata.csv", 
  mart = mart
)

wave2_counts <- get_summarized_experiment(
  htseq_counts, 
  "../data/w2_metadata.csv", 
  mart = mart
)

# prepare multiprocessing for RNA-seq differential expression
num_cores <- 5
param <- SnowParam(num_cores, "SOCK", progressbar = TRUE)
register(param)

```

## Load SOMA protein data

The SOMA data is loaded into long format and split into the "Wave 1" and 
"Wave 2" cohorts.

```{r load_prot_data, message=FALSE, warning=FALSE}

wave1_soma <- get_soma_data(
  soma_abundance = "../data/soma_abundance.csv", 
  sample_meta = "../data/w1_metadata.csv", 
  sample_technical_meta = "../data/sample_technical_meta.csv", 
  feature_meta = "../data/feature_meta.csv",
  ret_wide = FALSE
)

wave2_soma <- get_soma_data(
  soma_abundance = "../data/soma_abundance.csv", 
  sample_meta = "../data/w2_metadata.csv", 
  sample_technical_meta = "../data/sample_technical_meta.csv", 
  feature_meta = "../data/feature_meta.csv",
  ret_wide = FALSE
)

```

# Differential expression analyses of COVID-19 status

## RNA-seq analysis using dream and GSVA

### Wave 1 transcriptomic COVID-19 infected vs. non-infected comparison

```{r wave1_case_control_gene_de, message=FALSE, warning=FALSE, fig.show='hide'}

# apply normalisation to counts (returns edgeR DGE object)
wave1_case_control_dge <- normalize_se(
  wave1_counts,
  normalisation_level = "pre_dream",
  group = "case_control"
)

# run dream differential expression analysis (from variancePartition package)
wave1_case_control_de <- run_dream(
  f = ~ case_control + sex + ethnicity + calc_age + (1 | individual_id),
  dge = wave1_case_control_dge,
  L = list(case_control = "case_controlPOSITIVE"),
  L_is_contrasts = FALSE
)

write.csv(
    wave1_case_control_de$case_control_tt, 
    "../results/1_differential_expression/wave_1/gene_negative_vs_positive.csv", 
    row.names = FALSE
)

head(wave1_case_control_de$case_control_tt)

```

### Wave 2 transcriptomic COVID-19 infected vs. pre-infection comparison

```{r wave2_case_control_gene_de, message=FALSE, warning=FALSE, fig.show='hide'}

# apply normalisation to counts (returns edgeR DGE object)
# only NEGATIVE (pre-infection) and POSITIVE samples are compared in this analysis
wave2_case_control_dge <- normalize_se(
  wave2_counts[, wave2_counts$case_control != "RECOVERY"],
  normalisation_level = "pre_dream",
  group = "case_control"
)

# run dream differential expression analysis (from variancePartition package)
wave2_case_control_de <- run_dream(
  f = ~ case_control + sex + ethnicity + calc_age + (1 | individual_id),
  dge = wave2_case_control_dge,
  L = list(case_control = "case_controlPOSITIVE"),
  L_is_contrasts = FALSE
)

write.csv(
    wave2_case_control_de$case_control_tt, 
    "../results/1_differential_expression/wave_2/gene_negative_vs_positive.csv", 
    row.names = FALSE
)

print(head(wave2_case_control_de$case_control_tt))

```

### Wave 2 transcriptomic COVID-19 recovery vs. pre-infection comparison

```{r wave2_preinfection_recovery_gene_de, message=FALSE, warning=FALSE, fig.show='hide'}

# only NEGATIVE (pre-infection) and RECOVERY samples are compared in this analysis
wave2_recov_counts <- wave2_counts[, wave2_counts$case_control != "POSITIVE"]

# some patients do not have a RECOVERY sample
# since this is a paired analysis, these individual's pre-infection timepoints are removed
individuals_with_recovery_sample <- wave2_recov_counts$individual_id[wave2_recov_counts$case_control == "RECOVERY"]
wave2_recov_counts <- wave2_recov_counts[,  wave2_recov_counts$individual_id %in% individuals_with_recovery_sample]

# apply normalisation to counts (returns edgeR DGE object)
wave2_recovery_dge <- normalize_se(
  wave2_recov_counts, 
  normalisation_level = "pre_dream", 
  group = "case_control"
)

# run dream differential expression analysis (from variancePartition package)
wave2_recovery_de <- run_dream(
  f = ~ case_control + sex + ethnicity + calc_age + (1 | individual_id),
  dge = wave2_recovery_dge,
  L = list(case_control = "case_controlRECOVERY"),
  L_is_contrasts = FALSE
)

write.csv(
    wave2_recovery_de$case_control_tt, 
    "../results/1_differential_expression/wave_2/gene_negative_vs_recovery.csv", 
    row.names = FALSE
)

print(head(wave2_recovery_de$case_control_tt))

```

## Proteomic analysis using dream and GSVA

### Wave 1 proteomic COVID-19 infected vs. non-infected comparison

```{r wave1_case_control_prot_de, message=FALSE, warning=FALSE, fig.show='hide'}

wave1_case_control_models <- soma_de(
  soma_long = wave1_soma,
  formula_string = "RFU ~ case_control + sex + ethnicity + calc_age + (1 | individual_id)"
)

# takes a list of models and applies anova to each one
# the anova results are concatenated, written as a CSV, and returned as a data.frame
models_to_df <- function(m, file_path) {
    
    model_df <- data.frame()
    
    for (i in 1:length(m)) {
        
        anova_df <- data.frame(anova(m[[i]]))
        colnames(anova_df) <- c("sum_sq", "mean_sq", "num_df", "den_df", "f_value", "p_value")
        
        anova_df$feature <- names(m)[i]
        anova_df$term <- rownames(anova_df)

        model_df <- rbind(model_df, anova_df)
    }
    
    rownames(model_df) <- NULL
    model_df$adjusted_p_value <- NA
    
    for (term in unique(model_df$term)) {
        
        model_df$adjusted_p_value[model_df$term == term] <- p.adjust(model_df$p_value[model_df$term == term], method = "BH")
    }
    
    write.csv(model_df, file_path, row.names = FALSE)
    
    return(model_df)
}

wave1_case_control_df <- models_to_df(
    wave1_case_control_models, 
    "../results/1_differential_expression/wave_1/protein_negative_vs_positive.csv"
)

head(wave1_case_control_df[wave1_case_control_df$term == "case_control" ,])

```

### Wave 2 proteomic COVID-19 infected vs. non-infected comparison

```{r wave2_case_control_prot_de, message=FALSE, warning=FALSE, fig.show='hide'}

wave2_case_control_models <- soma_de(
  soma_long = wave2_soma,
  formula_string = "RFU ~ case_control + sex + ethnicity + calc_age + (1 | individual_id)"
)

wave2_case_control_df <- models_to_df(
    wave2_case_control_models, 
    "../results/1_differential_expression/wave_2/protein_negative_vs_positive.csv"
)

head(wave2_case_control_df[wave2_case_control_df$term == "case_control" ,])

```

# Differential expression analyses of COVID-19 severity at time of sampling

## RNA-seq

### Wave 1 transcriptomic COVID-19 severity analysis

```{r wave1_severity_gene_de, message=FALSE, warning=FALSE, fig.show='hide'}

wave1_counts$WHO_temp_severity <- ordered(wave1_counts$WHO_temp_severity, c("mild", "moderate", "severe", "critical"))

wave1_severity_dge <- normalize_se(
  wave1_counts[, wave1_counts$case_control == "POSITIVE" & !is.na(wave1_counts$WHO_temp_severity)],
  normalisation_level = "pre_dream",
  group = "WHO_temp_severity"
)

# run the differential expression
wave1_severity_de <- run_dream(
  f = ~ WHO_temp_severity + sex + ethnicity + calc_age + (1 | individual_id),
  dge = wave1_severity_dge,
  L = list("WHO_temp_severity" = c("WHO_temp_severity.L", "WHO_temp_severity.Q", "WHO_temp_severity.C")),
  L_is_contrasts = FALSE
)

write.csv(
    wave1_severity_de$WHO_temp_severity_tt, 
    "../results/1_differential_expression/wave_1/gene_severity.csv", 
    row.names = FALSE
)

print(head(wave1_severity_de$WHO_temp_severity_tt))

```

### Wave 2 transcriptomic COVID-19 severity analysis

```{r wave2_severity_gene_de, message=FALSE, warning=FALSE, fig.show='hide'}

wave2_counts$WHO_temp_severity <- ordered(wave2_counts$WHO_temp_severity, c("mild", "moderate", "severe", "critical"))

wave2_severity_dge <- normalize_se(
  wave2_counts[, wave2_counts$case_control == "POSITIVE" & !is.na(wave2_counts$WHO_temp_severity)],
  normalisation_level = "pre_dream",
  group = "WHO_temp_severity"
)

# run the differential expression
wave2_severity_de <- run_dream(
  f = ~ WHO_temp_severity + sex + ethnicity + calc_age + (1 | individual_id),
  dge = wave2_severity_dge,
  L = list(WHO_temp_severity = c("WHO_temp_severity.L", "WHO_temp_severity.Q", "WHO_temp_severity.C")),
  L_is_contrasts = FALSE
)

write.csv(
    wave2_severity_de$WHO_temp_severity_tt, 
    "../results/1_differential_expression/wave_2/gene_severity.csv", 
    row.names = FALSE
)

print(head(wave2_severity_de$WHO_temp_severity_tt))

```

## Proteomics

### Wave 1 proteomic COVID-19 severity analysis

```{r wave1_severity_prot_de, message=FALSE, warning=FALSE, fig.show='hide'}

wave1_soma$WHO_temp_severity <- ordered(wave1_soma$WHO_temp_severity, c("mild", "moderate", "severe", "critical"))
wave1_soma_positive <- wave1_soma[wave1_soma$case_control == "POSITIVE" & !is.na(wave1_soma$WHO_temp_severity) ,]

wave1_severity_models <- soma_de(
  soma_long = wave1_soma_positive,
  formula_string = "RFU ~ WHO_temp_severity + sex + ethnicity + calc_age + (1 | individual_id)"
)

wave1_severity_df <- models_to_df(
    wave1_severity_models, 
    "../results/1_differential_expression/wave_1/protein_severity.csv"
)

head(wave1_severity_df[wave1_severity_df$term == "WHO_temp_severity" ,])

```

### Wave 2 proteomic COVID-19 severity analysis

```{r wave2_severity_prot_de, message=FALSE, warning=FALSE, fig.show='hide'}

wave2_soma$WHO_temp_severity <- ordered(wave2_soma$WHO_temp_severity, c("mild", "moderate", "severe", "critical"))
wave2_soma_severity <- wave2_soma[wave2_soma$case_control == "POSITIVE" & !is.na(wave2_soma$WHO_temp_severity) ,]


wave2_severity_models <- soma_de(
  soma_long = wave2_soma_severity,
  formula_string = "RFU ~ WHO_temp_severity + sex + ethnicity + calc_age + (1 | individual_id)"
)

wave2_severity_df <- models_to_df(
    wave2_severity_models, 
    "../results/1_differential_expression/wave_2/protein_severity.csv"
)

head(wave2_severity_df[wave2_severity_df$term == "WHO_temp_severity" ,])

```

```{r session_info}
sessionInfo()
```

